---
title: "190536235_ST2195_Question2: Do older planes suffer more delays?"
output:
  html_document:
     df_print: paged
---
# Question 2: Do older planes suffer more delays? 
### Install and load the library

Install and load the `reshape2`` package.
```{r}
if (!("reshape2" %in% installed.packages())) {
  install.packages("reshape2")
}
library(reshape2)
```

Install and load the `dplyr`` package.
```{r}
if (!("dplyr" %in% installed.packages())) {
  install.packages("dplyr")
}
library(dplyr)
```

Install and load the `ggthemes`` package.
```{r}
if (!("ggthemes" %in% installed.packages())) {
  install.packages("ggthemes")
}
library(ggthemes)
```

Install and load the `ggplot2`` package.
```{r}
if (!("ggplot2" %in% installed.packages())) {
  install.packages("ggplot2")
}
library(ggplot2)
```

Install and load the `tidyr`` package.
```{r}
if (!("tidyr" %in% installed.packages())) {
  install.packages("tidyr")
}
library(tidyr)
```


### Data pre-processing stage (If data has not previously load from question 1.Rmd)
Load and save csv file where header is reflected. Note: results from analysis may slightly differ due to random subset samples.
```{r}
ontime_2008 <- read.csv("D:/SIM courses/Programming/Project/2008.csv", header = TRUE)
ontime_2007 <- read.csv("D:/SIM courses/Programming/Project/2007.csv", header = TRUE)
ontime2007_2008 <- rbind(ontime_2007, ontime_2008)

airports <- airports <-read.csv("D:/SIM courses/Programming/Project/airports.csv", header = TRUE)
carriers <- read.csv("D:/SIM courses/Programming/Project/carriers.csv", header = TRUE)
planes <- read.csv("D:/SIM courses/Programming/Project/plane-data.csv", header = TRUE)

```

Randomly subset 30% of the data set due to constraint on computer system, check and remove missing values. Additionally, remove unused data frames to free up spaces.    
```{r}
ontime <- ontime2007_2008 %>% sample_frac(0.3)
class(ontime)
sum(is.na(ontime))

rm(ontime_2007, ontime_2008, ontime2007_2008)
```

Assumption made on the criteria for outliers on departure and arrival delay to be removed when it exceeds 240 minutes (4hours) based on airline policies.
```{r}
df <- ontime %>%
  rename_all(tolower) %>%
  filter(depdelay<240 & arrdelay<240)
```
References: https://www.citizensadvice.org.uk/consumer/holiday-cancellations-and-compensation/if-your-flights-delayed-or-cancelled/


### Data Query
#### Question 2: Do older planes suffer more delays? 
Using issue dates as year of the plane that commercialized and used after manufacturing which then able to detect delays upon using.
```{r}
planes$issue_date <- as.Date(planes$issue_date,"%m/%d/%Y")
sum(is.na(planes$issue_date))
planes <- planes %>% 
  na_if("") %>%
  na.omit

```

Issue dates has been grouped into planes that commercialized Before and After 2000 due to technological advancement after 2000. The data has been filtered to only include non-cancelled flight and showing the average departure and arrival delays.
```{r}
q2 <- df %>% 
  rename_all(tolower) %>%
  inner_join(planes, by = "tailnum") %>%
  mutate(Period  = case_when((issue_date < "2000-01-01") ~ "Before 2000",
                             (issue_date > "2000-01-01") ~ "After 2000")) %>%
  select(depdelay, arrdelay, issue_date, tailnum, cancelled, Period) %>%
  filter(cancelled == 0) %>%
  group_by(issue_date, Period) %>%
  summarize(avg_depdelay = mean(depdelay, na.rm=TRUE),
            avg_arrdelay = mean(arrdelay, na.rm=TRUE)) %>%
  arrange(avg_depdelay)
q2

q2_melt <- melt(q2, id.var=c("issue_date", "Period"))
colnames(q2_melt) <- c("issue_date", "period", "delaytypes", "delayvalues")
q2_melt

```

Plot to show the average departure and arrival delays for planes commercialized Before and After 2000.
```{r}
q2_plot <- ggplot(q2_melt, aes(x=period, y = delayvalues, fill = delaytypes)) + 
  geom_violin(na.rm=TRUE, adjust=0.5) +
  stat_summary(fun = "mean", geom="text", size=4, position = position_dodge(width = 0.9), alpha=0.6, aes(label=round(..y.., digits=3)), hjust=-2, vjust = -2) +
  stat_summary(fun = "mean", geom="point", shape=23, size=2, position = position_dodge(width = 0.9), alpha=0.6) +
  ggtitle("Violinplot of Flight delays and Period of the Year") + 
  xlab("Period of the Year") + 
  ylab("Flights Delayed (in minutes)")
q2_plot 

```



References: 
Citizensadvice.org.uk. 2022. Claim compensation if your flight's delayed or cancelled. [online] Available at: <https://www.citizensadvice.org.uk/consumer/holiday-cancellations-and-compensation/if-your-flights-delayed-or-cancelled/> [Accessed 12 March 2022].