---
title: "190536235_ST2195_Question1: When is the best time of day, day of the week, and time of year to fly to minimise delays?"
output:
  html_document:
     df_print: paged
---
### Install and load the library

Install and load the `reshape2`` package.
```{r}
if (!("reshape2" %in% installed.packages())) {
  install.packages("reshape2")
}
library(reshape2)
```

Install and load the `dplyr`` package.
```{r}
if (!("dplyr" %in% installed.packages())) {
  install.packages("dplyr")
}
library(dplyr)
```

Install and load the `ggthemes`` package.
```{r}
if (!("ggthemes" %in% installed.packages())) {
  install.packages("ggthemes")
}
library(ggthemes)
```

Install and load the `ggplot2`` package.
```{r}
if (!("ggplot2" %in% installed.packages())) {
  install.packages("ggplot2")
}
library(ggplot2)
```

Install and load the `tidyr`` package.
```{r}
if (!("tidyr" %in% installed.packages())) {
  install.packages("tidyr")
}
library(tidyr)
```

### Data pre-processing stage
Load csv file where header row is reflected.
```{r}
ontime_2008 <- read.csv("D:/SIM courses/Programming/Project/2008.csv", header = TRUE)
ontime_2007 <- read.csv("D:/SIM courses/Programming/Project/2007.csv", header = TRUE)
ontime2007_2008 <- rbind(ontime_2007, ontime_2008)

airports <- airports <-read.csv("D:/SIM courses/Programming/Project/airports.csv", header = TRUE)
carriers <- read.csv("D:/SIM courses/Programming/Project/carriers.csv", header = TRUE)
planes <- read.csv("D:/SIM courses/Programming/Project/plane-data.csv", header = TRUE)
```

Randomly subset 30% of the data set due to constraint on computer system, check and remove missing values. Additionally, remove unused data frames to free up spaces for better loading performance. For lower-memory computer system, in order to run huge data to allocate vector size exceeding current computer performance, it is recommended to increase virtual memory size of at least two or three times of your current Random Access Memory (RAM) size. Note: results from analysis may slightly differ due to random subset samples.
```{r}
ontime <- ontime2007_2008 %>% sample_frac(0.3)
class(ontime)
sum(is.na(ontime))

rm(ontime_2007, ontime_2008, ontime2007_2008)
```

Assumption made on the criteria for outliers on departure and arrival delay to be removed when it exceeds 240 minutes (4hours) based on airline policies.
```{r}
df <- ontime %>%
  rename_all(tolower) %>%
  filter(depdelay<240 & arrdelay<240)
```
References: https://www.citizensadvice.org.uk/consumer/holiday-cancellations-and-compensation/if-your-flights-delayed-or-cancelled/


### Data Query
#### Question 1:  When is the best time of day, day of the week, and time of year to fly to minimise delays?

Grouping departure time into morning, afternoon and evening, day of week into respective name of weekdays and weekends and quarterly for over range of months, and calculate its average departure delay for non-cancelled flights
```{r}
q1 <- df %>%
  mutate(Day_Period  = case_when(between(deptime, 500, 1159) ~ "Morning",
                                 between(deptime, 1200, 1659) ~"Afternoon",
                                 between(deptime, 1700, 2059) ~"Evening",
                                 TRUE ~ "Night")) %>%
  mutate(NameOfDay  = case_when((dayofweek == 1) ~ "Mon",
                                (dayofweek == 2) ~ "Tues",
                                (dayofweek == 3) ~ "Wed",
                                (dayofweek == 4) ~ "Thurs",
                                (dayofweek == 5) ~ "Fri",
                                (dayofweek == 6) ~ "Sat",
                                TRUE ~ "Sun"))%>%
  mutate(Quarterly = case_when(between(month, 1,3) ~ "Q1",
                               between(month, 4,6) ~ "Q2",
                               between(month, 7,9) ~ "Q3",
                               TRUE ~ "Q4")) %>%
  rename_all(tolower) %>%
  select(day_period, nameofday, quarterly, cancelled, depdelay) %>%
  filter(cancelled == 0) %>%
  group_by( quarterly, nameofday, day_period) %>%
  summarize(avg_depdelay = mean(depdelay))%>%
  arrange(avg_depdelay)
```

Factor and re-level names for organization purposes.
```{r}
q1$day_period <- factor(q1$day_period, levels=c("Night", "Morning","Afternoon","Evening"))
levels(q1$day_period)
q1$nameofday <- factor(q1$nameofday, levels=c("Mon","Tues","Wed","Thurs","Fri","Sat", "Sun"))
levels(q1$nameofday)

```

Plot heatmap showing the time, day and period of the year to minimize flight departure delays.
```{r}
q1_heatmap <- ggplot(q1, aes(x = nameofday, 
                             y = day_period,
                             fill = avg_depdelay)) +
  geom_tile() +
  ggtitle("Average Departure delays of Day Time, Day Week and Quarterly") +
  xlab(label="Period of the day") +
  ylab(label="Day of Week") +
  facet_grid(~ quarterly,switch = "x",scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Avg_DepDelay",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside") +
  theme_bw()
q1_heatmap
```


References: 
Citizensadvice.org.uk. 2022. Claim compensation if your flight's delayed or cancelled. [online] Available at: <https://www.citizensadvice.org.uk/consumer/holiday-cancellations-and-compensation/if-your-flights-delayed-or-cancelled/> [Accessed 12 March 2022].