---
title: "190536235_ST2195_Question4: Can you detect cascading failures as delays in one airport create delays in others?"
output:
  html_document:
     df_print: paged
---
# Question 4: Can you detect cascading failures as delays in one airport create delays in others?

### Install and load the library

Install and load the `reshape2`` package.
```{r}
if (!("reshape2" %in% installed.packages())) {
  install.packages("reshape2")
}
library(reshape2)
```

Install and load the `dplyr`` package.
```{r}
if (!("dplyr" %in% installed.packages())) {
  install.packages("dplyr")
}
library(dplyr)
```

Install and load the `ggthemes`` package.
```{r}
if (!("ggthemes" %in% installed.packages())) {
  install.packages("ggthemes")
}
library(ggthemes)
```

Install and load the `ggplot2`` package.
```{r}
if (!("ggplot2" %in% installed.packages())) {
  install.packages("ggplot2")
}
library(ggplot2)
```

Install and load the `tidyr`` package.
```{r}
if (!("tidyr" %in% installed.packages())) {
  install.packages("tidyr")
}
library(tidyr)
```


### Data pre-processing stage (If data has not previously load from question 1.Rmd)
Load csv file where header is reflected. Note: results from analysis may slightly differ due to random subset samples.
```{r}
ontime_2008 <- read.csv("D:/SIM courses/Programming/Project/2008.csv", header = TRUE)
ontime_2007 <- read.csv("D:/SIM courses/Programming/Project/2007.csv", header = TRUE)
ontime2007_2008 <- rbind(ontime_2007, ontime_2008)

airports <- airports <-read.csv("D:/SIM courses/Programming/Project/airports.csv", header = TRUE)
carriers <- read.csv("D:/SIM courses/Programming/Project/carriers.csv", header = TRUE)
planes <- read.csv("D:/SIM courses/Programming/Project/plane-data.csv", header = TRUE)

```

Randomly subset 30% of the data set due to constraint on computer system, check and remove missing values. Additionally, remove unused data frames to free up spaces.    
```{r}
ontime <- ontime2007_2008 %>% sample_frac(0.3)
class(ontime)
sum(is.na(ontime))

rm(ontime_2007, ontime_2008, ontime2007_2008)
```

Assumption made on the criteria for outliers on departure and arrival delay to be removed when it exceeds 240 minutes (4hours) based on airline policies.
```{r}
df <- ontime %>%
  rename_all(tolower) %>%
  filter(depdelay<240 & arrdelay<240)
```
References: https://www.citizensadvice.org.uk/consumer/holiday-cancellations-and-compensation/if-your-flights-delayed-or-cancelled/



### Data Query
#### Question 4: Can you detect cascading failures as delays in one airport create delays in others?
To detect cascading failures of delays, data set has been filtered to only include non-cancelled flights that has departure delays more than 0minutes and calculates its average departure delays. 
```{r}
q4 <- df %>%
  rename_all(tolower) %>%
  select(origin, dest, depdelay, cancelled, dayofmonth, month, year, deptime, crsdeptime) %>% 
  filter(cancelled==0, depdelay >0) %>%
  group_by(dest ,origin, month, year, dayofmonth, deptime, crsdeptime) %>% 
  summarize(mean_depdelay = mean(depdelay))
q4
```

To analyze in-depth on cascading failures, data has been filtered to compare the average departure delays on 1st day of the month for 2007 and 2008.
```{r}
q4.1 <- q4 %>% filter(dayofmonth==1, month==1, year==2007) 
q4.1 <- q4.1 %>% group_by(dest) %>% top_n(1)
q4.1 <- head(q4.1,100, mean_depdelay)
q4.1

q4.2 <- q4 %>% filter(dayofmonth==1, month==1, year==2008) 
q4.2 <- q4.2 %>% group_by(dest) %>% top_n(1)
q4.2 <- head(q4.2,100,mean_depdelay)
q4.2

```

Install and load the `igraph`` package.
```{r}
if (!("igraph" %in% installed.packages())) {
  install.packages("igraph")
}
library(igraph)
```

Plot network graph to show cascading failures of departure delay affecting from origin to destination on the 1st day of the month.
```{r}

edgelist <- as.matrix(q4.1[c("origin", "dest")]) 
g_1 <- graph.edgelist(edgelist, direct=TRUE) 
E(g_1)
V(g_1)
g_1 <- simplify(g_1)
print(g_1, e=TRUE, v=TRUE)

plot.igraph(g_1, 
            edge.arrow.size=0.3,
            edge.curved=FALSE,
            edge.width=E(g_1)$mean_depdelay,
            edge.attribute=E(g_1)$mean_depdelay, 
            edge.color = "#FF847C",
            vertex.size=2,
            vertex.color=NA, 
            vertex.frame.color=NA, 
            vertex.label=V(g_1)$name,
            vertex.label.cex=1,
            layout=layout.circle,
            asp = 0.6,
            margin= -0.1
)
title("Departure Delay of Origin to Dest for 1st January 2007")
```

Plot network graph to show cascading failures of departure delay affecting from origin to destination airport on the 2nd day of the month.
```{r}
edgelist <- as.matrix(q4.2[c("origin", "dest")]) 
g_2 <- graph.edgelist(edgelist, direct=TRUE) 
E(g_2)
V(g_2)
g_2 <- simplify(g_2)
print(g_1, e=TRUE, v=TRUE)

plot.igraph(g_2, 
            edge.arrow.size=0.3,
            edge.curved=FALSE,
            edge.width=E(g_2)$mean_depdelay,
            edge.attribute=E(g_2)$mean_depdelay, 
            edge.color = "#FF847C",
            vertex.size=2,
            vertex.color=NA, 
            vertex.frame.color=NA, 
            vertex.label=V(g_2)$name,
            vertex.label.cex=1,
            layout=layout.circle,
            asp = 0.6,
            margin= -0.1
)
title("Departure Delay of Origin to Dest for 1st January 2008")
```


References: 
Citizensadvice.org.uk. 2022. Claim compensation if your flight's delayed or cancelled. [online] Available at: <https://www.citizensadvice.org.uk/consumer/holiday-cancellations-and-compensation/if-your-flights-delayed-or-cancelled/> [Accessed 12 March 2022]. 